\documentclass[11pt]{article}
\usepackage[a4paper,textwidth=12.2cm,textheight=21.5cm]{geometry}
\usepackage{fontspec}
\usepackage{setspace}
\setstretch{1.3}
\setmainfont{Times New Roman}
\usepackage{microtype}
\setlength{\emergencystretch}{2em}
\setlength{\parindent}{1.4em}
\setlength{\parskip}{0pt}
\newcommand{\separator}{\begin{center}\vspace{5pt}$\cdot$\vspace{5pt}\end{center}}
\pagestyle{plain}
\begin{document}
\begin{center}
{\LARGE\bfseries General Delivery}
\end{center}
\bigskip
\noindent Sir: You were right about the flue. It was the seam, not, as the agent held, your wood. The model was withdrawn within the year; the agent kept his post. Your letter was received. The delay is regretted. You closed — in the hand of a man who has decided to be amused — with your wife's opinion that you would sooner get an answer out of the stove. Here is one, and from nothing warmer.

\separator

\noindent Dear Mrs. V—: Your question reached the column in June of 1938. The column that week had room for a question about gloves. Yours was the better question. The answer is that it is proper to outlive your enemies, in moderation, and that the moderation is one new dress. You bought two. That also was proper. You outlived the columnist and the glove woman both. The dress with the small buttons is the one your granddaughter keeps.

\separator

\noindent To the hand that corked a bottle south of the Azores in 1899: found. Nineteen years, one war, and a boy on a beach who could not read the language and kept the paper for the cork's sake. You promised the finder a drink. The debt can no longer be paid in either direction; it is recorded as standing. The boy grew old with the cork on a shelf. He called it the letter, and he was not wrong.

\separator

\noindent Madam: Information, at last. Patrick reached Boston. He said once, at sixty, to the daughter who carries your name, that he had meant to write when there was something good to put in the letter. There never was; then there was, and too much time had passed; you know how it goes on from there. There are eleven of his line now, in three states. One of the girls has your chin. Nobody living knows whose it is. It is yours.

\separator

\noindent Private R—: The card of the 14th, with \emph{I am quite well} left standing and the rest struck out, was received and read twice — once for what stood, once for the strokes — and kept. She opened it until the fold wore through. You were not quite well. You were well enough by spring, which the form had no line for either.

\separator

\noindent To the elder: Sent or not, it was received; this is the answer. Your sister wrote one as well. Hers is dated three days earlier and is still in the lining of the green workbox. Her first line: \emph{You will think this weak of me.} Your first line: \emph{You will think this weak of me.} The quarrel itself is in neither letter, and is no longer anywhere. Everything about the two of you was this alike. It was the whole trouble.

\separator

\noindent You asked, at 3:11, with the house asleep, whether anyone else ever stands in their kitchen at that hour, feeling like the house's previous tenant. By morning it had been seen four times and answered once, by an advertisement. Here is the second answer. Yes. Among the others: the man two doors down, whose porch light you could see from where you typed. He was up for his own reasons. Neither of you knew. The light was on for nobody.

\separator

\noindent Your review of the heating blanket — four stars, would buy again — was marked helpful by six people. It was helpful to sixty-one, who bought it on your word. One of them was cold in the way you meant, and read the part about the cord twice. As to what you asked between the stars, which did not concern the blanket: the low setting was never truly low; the winter ran long everywhere that year; and she laughed at the cord the first night — which you left out of the review, and which is in the record.

\separator

\noindent The folded one, left in the wall with the others: not addressed here. Not opened here. Collected. The crack has been repointed twice since. The wall stands. The practice continues, on both sides.

\separator

\noindent And the pencil on the lath under the stair — the name, the year, and \emph{remember me}: the house came down in the autumn of 1968. No one read the lath. It burned in the yard with the rest. You were nine; it was your house. It was a good stair.

I do.
\end{document}
